Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-26 paper/paper.tex
Add more related work

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
The definition of in Figure~\ref{fig:logrelrec} explains how the $[[j
< i]]$ constraint in \rref{I-Set} makes our system predicative; the
interpretation of the $[[i]]_{th}$ universe is only dependent on
universes at strictly lower than $[[i]]$, which have been defined earlier.
Removing the ordering constraint would result in a
system where one can encode Girard's
paradox~\citep{girard-thesis}. Our proof would no longer work with the
restriction removed since the definition of $[[InterpR i A S]]$ would
not be well-founded when $[[Interp I i A S]]$ calls $[[I]]$ on
universe levels greater than or equal to $[[i]]$, which are yet to be defined.

@@

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Par} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Par} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
and \rref{I-Bool}.

@@

  \caption{Feature matrix for dependently typed languages with
    mechanized proofs by logical relation}
  \label{fig:linecount}
\end{figure}

\scw{Maybe it would be useful to include a chart here, so that readers can
  easily keep track of the features of the various languages.  i.e. which ones
  include large eliminations? type-directed equivalence? impredicative prop?
  inductive datatypes? what are their line counts?}

@@

\scw{Should give Martin L\"of credit for first consistency proof.}
\scw{What techniques do these three proofs use? How is it different from
your proof? What about Luo, which
introduced a universe hierarchy? What about categories with families, or
other ways of showing consistency?}

@@

More recent works such as \citet{Abel12}
and\citet{abel2008betaeta} make their definitions more explicit and
precise and thus easier to encode in proof assistants. Our definition
of logical relation is reminiscent of their definition of a
semantic universe hierarchy, though our logical relation is defined to
be closed under expansion with respect to parallel reduction rather
than weak head reduction. Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use the semantic universe hierachy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly because it is sufficient
for deriving consistency and later normalization.

\subsection{Mechanized logical relations about dependent types}

@@

>>>>

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I j]]
    := [[{A | exists S , InterpR j A S}]] \text{ for } [[j < i]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
The definition of in Figure~\ref{fig:logrelrec} explains how the $[[j
< i]]$ constraint in \rref{I-Set} makes our system predicative; the
interpretation of the $[[i]]_{th}$ universe is only dependent on
universes strictly lower than $[[i]]$, which have been defined earlier.
Removing the ordering constraint would result in a
system where one can encode Girard's
paradox~\citep{girard-thesis}. Our proof would no longer work with the
restriction removed since the definition of $[[InterpR i A S]]$ would
not be well-founded when $[[Interp I i A S]]$ calls $[[I]]$ on
universe levels greater than or equal to $[[i]]$, which are yet to be defined.

@@

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Red} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
and \rref{I-Bool}.

@@

  \caption{Feature matrix for dependently typed languages with
    mechanized proofs by logical relation}
  \label{fig:featurematrix}
\end{figure}

% \scw{Maybe it would be useful to include a chart here, so that readers can
%   easily keep track of the features of the various languages.  i.e. which ones
%   include large eliminations? type-directed equivalence? impredicative prop?
%   inductive datatypes? what are their line counts?}
% \yl{resolved}

@@

\scw{Should give Martin L\"of credit for first consistency proof.}
% \scw{What techniques do these three proofs use? How is it different from
% your proof? What about Luo, which
% introduced a universe hierarchy? What about categories with families, or
% other ways of showing consistency?}
% \yl{resolved}

@@

More recent works such as \citet{Abel12}
and\citet{abel2008betaeta} make their definitions more explicit and
precise and thus easier to encode in proof assistants. Our definition
of logical relation is reminiscent of their definition of a
semantic universe hierarchy, though our logical relation is defined to
be closed under expansion with respect to parallel reduction rather
than weak head reduction. Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use the semantic universe hierachy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly in our definition of
semantic typing because it is sufficient
for deriving consistency and later normalization.

\subsection{Mechanized logical relations about dependent types}
In Figure~\ref{fig:featurematrix}, we give an overview of features of
the object languages from mechanized logical relations about dependent
types. The table is far from exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.

\citet{casinghino:combining-proofs-programs} introduce $\lambda^\theta$, a dependently typed
programming language that uses modality to distinguish between logical
proofs and programs. The programmatic fragment
includes recursive data types and supports general recursion at the
cost of introducing divergence. Its logical fragment supports a
proof-irrelevant identity type, which can be used to reason about the
behavior of its programmatic fragment.
However, the $\lambda^\theta$
language is limited in its expressiveness since it does not support
type-level computation or polymorphism and therefore is not a fully dependently
typed language according to \citet{abel2013normalization}.
The consistency proof of $\lambda^\theta$'s logical fragment has been
mechanized in Coq through a step-indexed logical relation;
step-indexing is required to model the programmatic fragment, which
interacts with the logical fragment.
The lack of polymorphism and type-level computation means their
logical relation can be defined recursively for well-formed types using
a size metric, which has been explored in~\citet{liu2023dependently}.

@@

There are other mechanized developments for dependently typed systems
that only involve properties that are derivable through syntactic
means. For example, \citet{coqcoqcorrect2019} proves the correctness
of a type checker for the Polymorphic, Cumulative Calculus of
Inductive Constructions (PCUIC), Coq's core calculus, assuming the
strong normalization property of the object
language. \citet{weirich:systemd} define System D, a core calculus of
dependent Haskell, and prove the syntactic type soundness of the type
system. Compared to properties such as consistency, normalization, and
parametricity, syntactic type soundness is less intimidating for
dependent types since one is most unlikely to run into issues related
to the strength of the metatheory.
